---
title: 'Data / fig prep'
---

```{r}
#| label: setup
#| message: false
library(readxl)
library(tidyr)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(gt)
```

## EA-SIMD lookup table

A lookup table for mapping EA lables, to EA names, to SIMD ranking.
Requires reading SIMD table, and ToC from NPA data.
And some minor clean-up to allow joining


### SIMD

See [@mccartney2023]

SIMD overall education rank source:  
<https://datamap-scotland.co.uk/simd-local-authorities/>

```{r}
simd_path <- 'data/ea-rankings.csv'

simd_rankings <- read.csv(simd_path) |>
  rename(deprv_rank = deprv.rnk) |>
  mutate(
    local_authority = replace_values(
      local_authority,
      'Na h-Eileanan Siar' ~ 'Comhairle Nan Eilean Siar',
      'Glasgow City' ~ 'City Of Glasgow'
    )
  )
```


## NPA data

```{r}
npa_path <- 'data/npa.xlsx'
all_sheet_names <- excel_sheets(npa_path)
```

```{r}
#| label: create-ea-lookup-table
#| eval: false

ea_lookup <- read_xlsx(
  npa_path,
  sheet = 'Contents',
  n_max = 32,
  skip = 1
) |>
  rename(ea_name = 1) |>
  mutate(ea_name = sub('.* - ', '', ea_name)) |>
  mutate(
    ea_name = sub('[Tt]he ', '', ea_name),
    ea_name = sub(' +(and|&) +', ' And ', ea_name),
    ea_name = sub(' +of +', ' Of ', ea_name),
    ea_name = sub(' Council *', '', ea_name)
  ) |>
  mutate(ea_id = all_sheet_names[2:33], .before = 1)

# write.csv(ea_lookup, 'data/ea-lookup.csv', row.names = FALSE)
```

```{r}
ea_lookup_deprv <- left_join(
  ea_lookup,
  simd_rankings,
  by = join_by(ea_name == local_authority)
)
```

***********************************************************************

## Investigating NPAs

Steps:

Filter NPA data for only SCQF level 6.

Replace:

- confidential data `[c]` with NA
- Zero counts `[z]` with 0

```{r}
#| label: ea1-only
#| eval: false
npa_l6 <- read_xlsx(npa_path, sheet = 'EA1', skip = 3) |> #
  janitor::clean_names() |>
  filter(level == 'SCQF6') |>
  select(-level)

names(npa_l6)

npa_l6 |>
  pivot_longer(
    contains('count'),
    names_to = 'year',
    values_to = 'award_count',
    names_transform = ~ as.integer(gsub('\\D', '', .))
  ) |>
  mutate(
    award_count = replace_values(award_count, '[c]' ~ NA, '[z]' ~ '0'),
    award_count = as.integer(award_count),
    subject = sub('\\s+-\\s+SCQF6 *', '', subject)
  ) |>
  mutate(ea_id = 'EA1', .before = 1)
```

## Chose focus EAs

| EA ID | EA Name             | Deprivation ranking |
|-------|---------------------|----------|
| EA1   | Aberdeen City       | 16       |
| EA2   | Aberdeenshire       | 30       |
| EA5   | City Of Glasgow     |  2       |
| EA29  | City Of Edinburgh   | 19       |
| EA11  | East Dunbartonshire | 24       |
| EA31  | West Dunbartonshire |  1       |

Compare Aberdeen City with Aberdeenshire (mid vs high SIMD)
Compare Glasgow City with Edinburgh (low vs mid SIMD)
Compare East vs West Dunbartonshire (high vs low SIMD)

NB: low rankings are most deprived.

```{r}
#| label: final-lookup-table
chosen_eas <- filter(
  ea_lookup_deprv,
  deprv_rank %in% c(1, 2, 16, 19, 24, 30)
) |>
  mutate(
    ea_rank = paste0(
      c(
        'Aberdeen City',
        'Aberdeenshire',
        'Glasgow',
        'E Dunbarton',
        'Edinburgh',
        'W Dunbarton'
      ),
      ' (',
      deprv_rank,
      ')'
    ),
    ea_rank = factor(ea_rank, ea_rank[order(chosen_eas$ea_name)])
  ) |>
  mutate(
    facet_var = case_when(
      grepl('Aber', ea_name) ~ 'Aberdeen',
      grepl('City', ea_name) ~ 'Ed-Gla',
      grepl('Dunb', ea_name) ~ 'Dunbarton'
    )
  )

sort(chosen_eas$ea_name)

chosen_eas$ea_rank[order(unique(chosen_eas$ea_name))]
chosen_eas$ea_rank |> levels()
```

### Make final analysis data

Pull data form all tables.

```{r}
#| label: data-prep-function

process_npa_data <- function(data_path, sheet) {
  read_xlsx(data_path, sheet = sheet, skip = 3) |>
    janitor::clean_names() |>
    filter(level == 'SCQF6') |>
    filter_out(if_all(contains('count'), ~ is.na(.) | . == 0)) |>
    select(-level) |>
    mutate(subject = sub(' - .*', '', subject)) |>
    pivot_longer(
      contains('count'),
      names_to = 'year',
      values_to = 'award_count',
      names_transform = ~ as.integer(gsub('\\D', '', .))
    ) |>

    mutate(
      award_count = replace_values(award_count, '[c]' ~ NA, '[z]' ~ '0'),
      award_count = as.integer(award_count),
      subject = sub('\\s+-\\s+SCQF6 *', '', subject)
    ) |>
    mutate(ea_id = sheet, .before = 1)
}
```

```{r}
#| label: prep-data
all_subj <- map(chosen_eas$ea_id, \(sheet) {
  process_npa_data(npa_path, sheet)
}) |>
  list_rbind()
```


```{r}
high_vol_awards <- all_subj |>
  filter(subject != 'Total', year == max(year)) |>
  summarize(total_n = sum(award_count, na.rm = T), .by = c(subject)) |>
  arrange(desc(total_n)) |>
  filter(total_n >= 100)
```

```{r}
#| warning: false
all_subj |>
  filter(!is.na(award_count)) |>
  inner_join(high_vol_awards, by = join_by(subject == subject)) |>
  inner_join(chosen_eas, by = join_by(ea_id == ea_id)) |>

  ggplot(aes(
    year,
    award_count,
    colour = ea_rank,
    group = paste0(ea_id, subject)
  )) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(subject)) +
  scale_color_brewer(type = 'qual', palette = 'Paired')
```

```{r}
all_subj |>
  filter(!is.na(award_count)) |>
  inner_join(filter(high_vol_awards, total_n %in% c(480, 370)), , by = join_by(subject == subject)) |>
  inner_join(chosen_eas, by = join_by(ea_id == ea_id)) |>
  mutate(subject = case_when(
    substr(subject, 1, 1) == 'B' ~ 'Business with IT',
    substr(subject, 1, 1) == 'E' ~ 'Fitness Leadership',
    .default = subject
  )) |> 
  ggplot(aes(
    year,
    award_count,
    colour = ea_rank,
    group = paste0(ea_id, subject)
  )) +
  geom_point(size = 3) +
  geom_line(linewidth = 1.2) +
  facet_wrap(vars(subject)) +
  scale_color_brewer(type = 'qual', palette = 'Paired') +
  labs(x = '', y = 'Award count', colour = '') +
  theme(text = element_text(size = 14)) +
  theme_jw_default()

# ggsave('figs/fig-award-counts.png')
```


Business with IT growth is being driven by Glasgow and E Dunb
W Dunb is near bottom for everything


```{r}
subject_summary <- all_subj |>
  inner_join(high_vol_awards) |>
  filter(year == 2025) |>
  summarize(
    total_awards = sum(award_count, na.rm = TRUE),
    mean_award = mean(award_count, na.rm = TRUE),
    sd_award = sd(award_count, na.rm = TRUE),
    iqr = IQR(award_count, na.rm = TRUE), #sd_award / mean_award,
    .by = subject
  ) |> 
  inner_join(high_vol_awards) |>
  select(subject, total_awards, iqr) |> 
  arrange(desc(total_awards))

```


```{r}
chosen <- c(
  'Exercise and Fitness Leadership',
  'Business with Information Technology',
  'Criminology'
)

tab_1 <- subject_summary %>%
  gt() %>%
  tab_header(
    title = 'NPA SCQF Level 6 Subject Selection',
    subtitle = 'Total awards for 2025 across 6 selected Education Authorities'
  ) %>%
  cols_label(
    subject = 'Subject',
    total_awards = 'Total Awards',
    iqr = 'IQR'
  ) %>%
  fmt_number(
    columns = iqr,
    decimals = 2
  ) %>%
  fmt_number(
    columns = total_awards,
    decimals = 0,
    use_seps = TRUE
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = '#2297E6'),
      cell_text(color = 'white', weight = 'bold')
    ),
    locations = cells_body(
      rows = subject %in% chosen
    )
  ) %>%
  tab_style(
    style = cell_text(weight = 'bold'),
    locations = cells_column_labels()
  ) %>%
  tab_footnote(
    footnote = 'Sports Development passed over in favour of topic breadth.',
    locations = cells_body(columns = total_awards, rows = 3),
    placement = 'right'
  ) %>%
  tab_options(
    table.font.size = 16,
    heading.title.font.size = 14,
    heading.subtitle.font.size = 11
  )


# gtsave(tab_1, filename = 'figs/tab_subject-selection.html')
```


```{r}
all_subj |>
  filter(year == 2025) |> 
  inner_join(high_vol_awards) |>
  summarize(
    total_awards = sum(award_count, na.rm = TRUE),
    iqr = IQR(award_count, na.rm = TRUE),
    .by = subject
  ) |>
  mutate(rank_iqr = rank(iqr)) |> 
  arrange(desc(rank_iqr)) |> 
  ggplot(aes(iqr, reorder(subject, iqr), fill = rank_iqr %in% c(5,7,9))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(1.3, subject, label = round(iqr, 3)), hjust = 0) +
  scale_fill_manual(breaks = c(TRUE, FALSE), values = c(4, 8)) +
  labs(title = 'Within subject variantion', y = '')

# ggsave(filename = 'figs/fig_subject-selection.png')
```

## Three chosen subjects- 

- Business with IT
- Criminology
- Exercise & Fitness Leadership

High volume (> 100 total awards in 2025)
High between-EA variance (IQR > 1.0)

Narrative

- one growth/digital subject (Business with IT)
- one social science (Criminology)
- one health/active living subject (Exercise & Fitness Leadership)



## Business with Information Technology

Glasgow dominates at 180 in 2025, up from just 15 in 2019.
East Dunbartonshire (125) and Aberdeen City (65) follow.
Edinburgh is surprisingly low at 55 despite being a large city.
West Dunbartonshire is near zero throughout - only 10 in 2025 with most years suppressed or zero.

Strong deprivation story: most deprived EA barely participates in the fast-growing subject.


## Criminology

Completely absent from Edinburgh across all 7 years - every single row is zero.
Surprising for capital.
Aberdeenshire and Glasgow lead at 125, East Dunbartonshire at 75.
West Dunbartonshire shows modest but consistent participation (30–40) from 2022 onwards, which is actually one of its stronger subjects relative to others.


## Exercise and Fitness Leadership

Edinburgh completely dominates at 185 in 2025 - nearly double anyone else.
Glasgow (90) and Aberdeen City (80) are next closest.
West Dunbartonshire is consistently very low (max 15), and had zero 2019–2021.


## Narrative

West Dunbartonshire (rank 1, most deprived) shows consistently low participation across almost every subject, while less deprived authorities grow strongly.
But there are interesting nuances.
Edinburgh's complete absence from Criminology and Glasgow's strength in Business IT complicate the simple narrative that deprivation = low participation.



```{r}
#| label: 'EA names are followed by brackets containing the EAs SIMD ranking. Note figure shows raw counts.'

total_award_data <- all_subj |>
  filter(subject == 'Total', !is.na(award_count)) |>
  inner_join(chosen_eas, by = join_by(ea_id))

ea_anno <- filter(total_award_data, subject == 'Total', year == 2025)


total_award_data |>
  ggplot(aes(year, award_count, colour = ea_rank)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 4) +
  labs(colour = 'Education Authority') +
  geom_text(
    aes(2025, award_count, label = ea_rank),
    filter(ea_anno, year == 2025),
    size = 6,
    hjust = 0,
    nudge_x = .2
  ) +
  theme_jw_default() +
  theme(
    # text = element_text(size = 22),
    #  legend.position = c(.2, 0.8),
    #  legend.background = element_rect(fill = 'white', color = 'black')
  ) +
  labs(x = '', y = 'Award count') +
  guides(colour = 'none') +
  scale_x_continuous(limits = c(2019, 2027.5), breaks = 2019:2025) +
  scale_colour_brewer(type = 'qual', palette = 'Paired')
  # labs(title = 'Total award counts from 2019 to 2025')

# ggsave('figs/fig_total-awards.png')
```


