---
title: "Untitled"
format: html
---

```{r}
#| label: setup
LoadPackages(rvest, httr2, readxl, tidyr, dplyr, stringr, purrr, ggplot2)
```

```{r}
#| label: setup-rvest-session 
u <- 'https://www.sqa.org.uk/sqa/'
s <- rvest::session(u)

att_stat_data_paths <- s |>
  rvest::session_jump_to('105123.11712.html') |>
  rvest::html_elements('a') |>
  rvest::html_attr('href') |>
  grepv('xlsx$', x = _)

att_stat_2025 <- grepv('2025', att_stat_data_paths)
sub('.*\\d\\d\\d\\d-', '', att_stat_2025)
```

```{r}
#| labal: dl-data
data_to_dl <- file.path(u, att_stat_2025[[33]])
save_to <- 'npa.xlsx'

if (!file.exists(save_to)) {
  download.file(data_to_dl, save_to, mode = 'wb')
}
```

```{r}
#| label: get-toc-from-first-sheet
nat_5_toc <- read_xlsx(
  'nat_5.xlsx',
  sheet = 'Contents',
  range = 'A2:A34'
) |>
  rename(toc = 1)
```

```{r}
#| label: extract-all-ea-names
ea_lookup <- nat_5_toc |>
  separate_wider_regex(
    cols = 'toc',
    patterns = c('.+ ', table = '\\d+', ':.+ - ', ea_name = '.+')
  ) |>
  mutate(
    table = paste0('EA', table),
    .after = 1,
    ea_name = sub('[Tt]he ', '', ea_name),
    ea_name = sub(' +(and|&) +', ' And ', ea_name),
    ea_name = sub(' +of +', ' Of ', ea_name),
    ea_name = sub(' Council *', '', ea_name)
  )
```

```{r}
#| label: combine-all-ea-data
all_nat5_data <- map(
  ea_lookup$table[1:2], # Just EA1 & EA2 for now
  ~ {
    read_xlsx('nat_5.xlsx', sheet = .x, skip = 3) |>
      select(Subject, contains('Percent')) |>
      mutate(
        across(everything(), ~ trimws(sub('%|,', '', .))),
        across(-Subject, ~ replace_values(., c('[c]', '[z]', '') ~ NA)),
        across(-Subject, as.numeric)
      )
  }
) |>
  list_rbind()

# Still just EA1 & EA2 for now
all_nat5_data

object.size(all_nat5_data)
```


```{r}
#| label: subject-counts
subjects_counts_2025 <- all_nat5_data |>
  mutate(
    across(contains('Percent'), ~ sub('%', '', .)),
    across(-1, as.numeric)
  ) |>
  select(1, contains('Count')) |>
  rowwise() |>
  transmute(
    Subject,
    total_count = sum(c_across(2:36), na.rm = TRUE)
  ) |>
  arrange(desc(total_count))

reframe(Subject, across(where(is.numeric), ~ min(., na.rm = T)))
```

__Idea:__ choose a small number of subjects, and compare across EAs?

***********************************************************************

## Investigating NPAs

```{r}
excel_sheets('npa.xlsx')
```

```{r}
npa_toc <- read_xlsx('npa.xlsx', skip = 1)
```

Confirmed all npa_ea == ea_lookup$desc

```{r}
npa_l6 <- read_xlsx('npa.xlsx', sheet = 'EA1', skip = 3) |> #
  janitor::clean_names() |>
  filter(level == 'SCQF6')

names(npa_l6)

npa_l6 |>
  pivot_longer(
    -(1:2),
    names_to = 'year',
    values_to = 'award_count',
    names_transform = ~ as.integer(gsub('\\D', '', .))
  ) |>
  mutate(
    award_count = replace_values(award_count, '[c]' ~ NA, '[z]' ~ '0'),
    award_count = as.integer(award_count),
    subject = sub('\\s+-\\s+SCQF6 *', '', subject)
  ) |>
  mutate(ea = 'EA1', .before = 1)
#   c('[c]', '[z]'), c(NA, 0)))
```



## SIMD

See [@mccartney2023]


```{r}
excel_sheets('../../SIMD+2020v2+-+ranks.xlsx')
read_xlsx('../../SIMD+2020v2+-+ranks.xlsx', sheet = 2)
```


```{r}
# overall rank source: https://datamap-scotland.co.uk/simd-local-authorities/
simd_ed_rank <- read.csv('../../data/ed-auth-ranking.csv') |>
  mutate(
    local_authority = replace_values(
      local_authority,
      'Na h-Eileanan Siar' ~ 'Comhairle Nan Eilean Siar',
      'Glasgow City' ~ 'City Of Glasgow'
    )
  )

ea_lookup_deprv <- left_join(
  ea_lookup,
  simd_ed_rank,
  by = join_by(ea_name == local_authority)
)
```

```{r}
npa_l6 |>
  pivot_longer(
    -(1:2),
    names_to = 'year',
    values_to = 'award_count',
    names_transform = ~ as.integer(gsub('\\D', '', .))
  ) |>
  mutate(
    award_count = replace_values(award_count, '[c]' ~ NA, '[z]' ~ '0'),
    award_count = as.integer(award_count),
    subject = sub('\\s+-\\s+SCQF6 *', '', subject)
  ) |>
  mutate(ea = 'EA1', .before = 1)
```

```{r}
npa_l6[-(1:2)] |>
  mutate(across(everything(), as.numeric)) |>
  filter_out(if_all(everything(), ~ is.na(.) | . == 0))

# only 2 rows have no 0s or NAs
npa_l6[-(1:2)] |>
  mutate(across(everything(), as.numeric)) |>
  filter_out(if_any(everything(), ~ is.na(.) | . == 0))

# 0 rows are all na
npa_l6[-(1:2)] |>
  mutate(across(everything(), as.numeric)) |>
  filter(if_a(everything(), is.na))

# 28 rows are all 0
npa_l6[-(1:2)] |>
  mutate(across(everything(), as.numeric)) |>
  filter(if_all(everything(), ~ . == 0))
```

```{r}

```

```{r}
abbrev_ea <- inner_join(
  simd_ed_rank,
  abbrev_ea,
  by = join_by(local_authority == ea_name)
)
```


```{r}
chosen_eas <- filter(ea_lookup_deprv, deprv_rank %in% c(1, 2, 16, 19, 24, 30))


all_subj <- map(chosen_eas$table, \(s) {
  read_xlsx('npa.xlsx', sheet = s, skip = 3) |>
    janitor::clean_names() |>
    filter(level == 'SCQF6') |>

    filter_out(if_all(-(1:2), ~ is.na(.) | . == 0)) |>

    select(-level) |>
    mutate(subject = sub(' - .*', '', subject)) |>
    pivot_longer(
      contains('count'),
      names_to = 'year',
      values_to = 'award_count',
      names_transform = ~ as.integer(gsub('\\D', '', .))
    ) |>

    mutate(
      award_count = replace_values(award_count, '[c]' ~ NA, '[z]' ~ '0'),
      award_count = as.integer(award_count),
      subject = sub('\\s+-\\s+SCQF6 *', '', subject)
    ) |>
    mutate(ea = s, .before = 1)
}) |>
  list_rbind()


abbrev_ea <- inner_join(ea_lookup, all_subj, by = join_by(table == ea)) |>
  distinct(table, ea_name) |>
  inner_join(simd_ed_rank, by = join_by(ea_name == local_authority)) |>
  mutate(
    ea_abbrev = paste0(
      c(
        'Aberdeen City',
        'Aberdeenshire',
        'Glasgow',
        'E Dumbarton',
        'Edinburgh',
        'W Dumbarton'
      ),
      ' (',
      deprv_rank,
      ')'
    )
  ) |>
  arrange(ea_name)
```

```{r}
all_subj |>
  filter_out(subject == 'Total') |>
  filter(year == max(year)) |>
  group_by(subject) |>
  summarise(n = n())

all_subj |>
  filter_out(subject == 'Total') |>
  filter(year == max(year)) |>
  count(subject, sort = TRUE)

all_subj |>
  inner_join(all_occur, by = join_by(subject == subject)) |>
  summarize(t = sum(award_count, na.rm = T), .by = c(subject, ea)) |>
  pivot_wider(names_from = ea, values_from = t)

all_occur <- all_subj |>
  filter(subject != 'Total', year == max(year)) |>
  summarize(total_n = sum(award_count, na.rm = T), .by = c(subject)) |>
  arrange(desc(total_n)) |>
  filter(total_n >= 100)
```

```{r}
#| warning: false
all_subj |>
  filter(!is.na(award_count)) |>
  inner_join(all_occur, by = join_by(subject == subject)) |>
  inner_join(abbrev_ea, by = join_by(ea == table)) |>
  mutate(
    comp = recode_values(
      ea,
      c('EA1', 'EA2') ~ 'Aberdeen',
      c('EA5', 'EA29') ~ 'Glw/Edb',
      default = 'Dumbarton'
    )
  ) |>
  ggplot(aes(
    year,
    award_count,
    colour = ea_abbrev,
    group = paste0(ea, subject)
  )) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(subject))
```

Business with IT growth is being driven by Glasgow and E Dumb

```{r}
all_subj |>
  filter(grepl('Legal', subject))
```

```{r}
all_subj |>
  inner_join(all_occur) |>
  summarize(
    cv = round(
      sd(award_count, na.rm = TRUE) / mean(award_count, na.rm = TRUE),
      3
    ),
    .by = subject
  ) |>
  arrange(desc(cv)) |>
  simplermarkdown::md_table(as_character = TRUE) |>
  writeClipboard()
```

## Three chosen subjects- 

- Business with IT
- Criminology
- Exercise & Fitness Leadership

High volume (>100 total awards in 2025)
High between-EA variance (CV > 1.0)
Narrative
    - one growth/digital subject (Business with IT)
    - one emerging social science (Criminology)
    - one health/active living subject (Exercise & Fitness Leadership)



## Business with Information Technology

Glasgow dominates at 180 in 2025, up from just 15 in 2019.
East Dunbartonshire (125) and Aberdeen City (65) follow.
Edinburgh is surprisingly low at 55 despite being a large city.
West Dunbartonshire is near zero throughout - only 10 in 2025 with most years suppressed or zero.

Strong deprivation story: most deprived EA barely participates in the fast-growing subject.


## Criminology

Completely absent from Edinburgh across all 7 years - every single row is zero.
Surprising for capital.
Aberdeenshire and Glasgow lead at 125, East Dunbartonshire at 75.
West Dunbartonshire shows modest but consistent participation (30–40) from 2022 onwards, which is actually one of its stronger subjects relative to others.


## Exercise and Fitness Leadership

Edinburgh completely dominates at 185 in 2025 - nearly double anyone else.
Glasgow (90) and Aberdeen City (80) are next closest.
West Dunbartonshire is consistently very low (max 15), and had zero 2019–2021.


## Clear story

West Dunbartonshire (rank 1, most deprived) shows consistently low participation across almost every subject, while less deprived authorities grow strongly.
But there are interesting nuances.
Edinburgh's complete absence from Criminology and Glasgow's strength in Business IT complicate the simple narrative that deprivation = low participation.



```{r}
all_subj |>
  filter(subject == 'Total') |>
  inner_join(abbrev_ea, by = join_by(ea == table)) |>
  #select(-subject) |>
  #  pivot_wider(names_from = year, values_from = award_count) |>
  ggplot(aes(year, award_count, colour = ea_abbrev)) +
  geom_line() +
  geom_point()
```


```{r}

```